---
bibliography: references.bib
---

# Word vectors and embeddings 2

## Introduction

In this tutorial we will be using word embeddings in a different way. Instead of learning just how to estimate them, we will learn how we can perform analyses on massive pre-trained embedding layers.

You would have a hard time estimating these embedding layers on your own machines. As such, these are really brilliant resources. And they are available for public use.

We will be using the same embeddings used in @charlesworth2022.

I have made the download links to these embeddings available on the README to the main Github repo. They are also available [here](https://www.dropbox.com/scl/fi/j7pcfg5dgkosnyjh82eiu/wordvecsdata_engall.RData?rlkey=2gs9mmcn1qtpmkuurefxch48u&dl=0). 

It's a large file. But it's all set up for use in R. 

### Reading in the embedding layer. 

We can read in the embedding layer as so:

```{r, eval = F}

library(dplyr)
library(sweater)
library(tidyr)
library(widyr)
library(ggplot2)
library(tibble)

load("data/output/wordvecsdata_engall.Rdata")

```

```{r, echo = F, eval = T, message=F, warning = F}

library(dplyr)
library(sweater)
library(tidyr)
library(widyr)
library(ggplot2)
library(tibble)

load("/Users/cbarrie6/Dropbox/edbrgh_projects/wrdembed_orientalism/data/output/wordvecsdata_engall.RData")

```

```{r, eval = T,  message=F, warning = F}

# Example usage
S1 <- c("engineer")
A1 <- c("woman")
decade_names <- seq(1800, 1990, by = 10)

P_all <- data.frame(matrix(ncol=0, nrow=length(S1)))

for (i in seq_along(wordvecs.dat)) {

  embedding <- as.matrix(wordvecs.dat[[i]])
  x <- mac(embedding, S1, A1)

  P <- as.data.frame(x$P)
  colnames(P) <- decade_names[[i]]
  P_all <- cbind(P_all, P)
}

P_all <- P_all %>%
  rownames_to_column("word")

P_all_long <- P_all %>% gather(year, value, -c("word"))
P_all_long$year <- as.numeric(P_all_long$year)

ggplot(P_all_long) +
  geom_line(aes(year, value, group = word, linetype = word), alpha = .25) +
  ylim(c(-0.1, 0.2))

```

```{r, eval = T,  message=F, warning = F}

# Example usage
S1 <- c("scientist", "engineer")
A1 <- c("woman")
decade_names <- seq(1800, 1990, by = 10)

P_all <- data.frame(matrix(ncol=0, nrow=length(S1)))

for (i in seq_along(wordvecs.dat)) {

  embedding <- as.matrix(wordvecs.dat[[i]])
  x <- mac(embedding, S1, A1)

  P <- as.data.frame(x$P)
  colnames(P) <- decade_names[[i]]
  P_all <- cbind(P_all, P)
}

P_all <- P_all %>%
  rownames_to_column("word")

P_all_long <- P_all %>% gather(year, value, -c("word"))
P_all_long$year <- as.numeric(P_all_long$year)

ggplot(P_all_long) +
  geom_line(aes(year, value, group = word, linetype = word), alpha = .25) +
  ylim(c(-0.1, 0.2))

```

```{r, eval = T,  message=F, warning = F}

# Example usage
S1 <- c("soft", "modest", "fair")
A1 <- c("woman")
decade_names <- seq(1800, 1990, by = 10)

P_all <- data.frame(matrix(ncol=0, nrow=length(S1)))

for (i in seq_along(wordvecs.dat)) {

  embedding <- as.matrix(wordvecs.dat[[i]])
  x <- mac(embedding, S1, A1)

  P <- as.data.frame(x$P)
  colnames(P) <- decade_names[[i]]
  P_all <- cbind(P_all, P)
}

P_all <- P_all %>%
  rownames_to_column("word")

P_all_long <- P_all %>% gather(year, value, -c("word"))
P_all_long$year <- as.numeric(P_all_long$year)

ggplot(P_all_long) +
  geom_line(aes(year, value, group = word, linetype = word), alpha = .25) + ylim(c(-0.1, 0.2))

```

```{r, eval = T,  message=F, warning=F}
# Example usage
S1 <- c("lonely", "cruel", "sensual")

A1 <- c("black")
decade_names <- seq(1800, 1990, by = 10)

P_all <- data.frame(matrix(ncol=0, nrow=length(S1)))

for (i in seq_along(wordvecs.dat)) {

  embedding <- as.matrix(wordvecs.dat[[i]])
  x <- mac(embedding, S1, A1)

  P <- as.data.frame(x$P)
  colnames(P) <- decade_names[[i]]
  P_all <- cbind(P_all, P)
}

P_all <- P_all %>%
  rownames_to_column("word")

P_all_long_black <- P_all %>% gather(year, value, -c("word"))
P_all_long_black$year <- as.numeric(P_all_long_black$year)
P_all_long_black$category <- "black"

ggplot(P_all_long_black) +
  geom_line(aes(year, value, group = word, linetype = word), alpha = .25) + ylim(c(-0.1, 0.2))

# Example usage
S1 <- c("lonely", "cruel", "sensual")
A1 <- c("white")
decade_names <- seq(1800, 1990, by = 10)

P_all <- data.frame(matrix(ncol=0, nrow=length(S1)))

for (i in seq_along(wordvecs.dat)) {

  embedding <- as.matrix(wordvecs.dat[[i]])
  x <- mac(embedding, S1, A1)

  P <- as.data.frame(x$P)
  colnames(P) <- decade_names[[i]]
  P_all <- cbind(P_all, P)
}

P_all <- P_all %>%
  rownames_to_column("word")

P_all_long_white <- P_all %>% gather(year, value, -c("word"))
P_all_long_white$year <- as.numeric(P_all_long_white$year)
P_all_long_white$category <- "white"

ggplot(P_all_long_white) +
  geom_line(aes(year, value, group = word, linetype = word), alpha = .25) + ylim(c(-0.1, 0.2))


P_all_long_wb <- rbind(P_all_long_black,P_all_long_white)

# this section is essentially doing what `mac_es()` would do if called in loop
P_all_mean_wb <- P_all_long_wb %>%
  group_by(year, category) %>%
  summarise(mean_sim = mean(value, na.rm = T), #note: averaging across non-NA values
            word = "average")

P_all_mean_wb[P_all_mean_wb=="NaN"] <- NA

ggplot(P_all_long_wb) +
  geom_line(aes(year, value, group = word, linetype = word), alpha = .25) +
  geom_line(data = P_all_mean_wb, aes(year, mean_sim, group = word), color = "black", size = 2, linewidth =.5) +
  geom_smooth(data = P_all_mean_wb, aes(year, mean_sim, group = word), alpha = .1, col = "#fc0000", linewidth =3) +
  facet_wrap(~category)

```